An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-06-07 cal_demo/ms.tex
Addressing more small reviewer comments

Images created at the beginning of calibration and at the end are shown in 
figure~\ref{fig:sim_images}. Each image is obtained over 10~ms integration, corresponding to 
all snapshot images created with a given set of gain estimates. The top panel shows the image 
produced with our initialized unity gains. Because the phases are completely random, the 
image is essentially noise with the primary beam evident. After 20 iterations, the image is far 
more clear, shown in the bottom panel. Each of the ten simulated sources are clearly visible, 
indicated with red circles.

@@

\section{Application to LWA data}\label{sec:data}
We next demonstrate our calibration algorithm using an observation from the LWA station in 
New Mexico. The data is from the LWA narrow-band transient buffer (TBN), with time ordered 
voltage data from 255 antennas within a core radius of 100~m. The central frequency is 74.03 
MHz, with a bandwidth of 100 kHz and writeout timescale of 5.12 ms (frequency channel 
resolution of 195.3125 Hz). For this demonstration we limit ourselves to a single polarization.

@@

We made several choices while studying the behavior of the LWA data to improve our 
calibration. Through our previous imaging work, we noted that the flux scale of uncalibrated 
images was consistent with average gain amplitudes of 0.25. To allow the calibration to 
converge quickly we initialized our gain estimates at this level. We also found a boost in signal 
to noise is achieved easily by averaging frequency channels and assuming the gains are 
constant within a sub-band. Here we average solutions across 150 channels, or about 29 kHz. 
With a fractional bandwidth $B/f_0 = 3.9 \times 10^{-4} \ll 1$, we assume a smooth bandpass 
across the band. 
A damping factor of $\gamma = 0.7$ was adopted.

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, 
and 244 were omitted.} produced unstable gain solutions, and in fact corrupted the entire array. 
We therefore omitted these antennas from our analysis, resulting in a total of 248 antennas to 
calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and 
update our gain estimates over 10 timestamps (51.2 ms). We iterate the loop 30 times for a 
total of 1.536 seconds. The results of this calibration experiment are shown in figures~
\ref{fig:data_phase}~--~\ref{fig:data_images}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.3\linewidth]{fig8a.pdf}
\includegraphics[width=0.3\linewidth]{fig8b.pdf}
\includegraphics[width=0.3\linewidth]{fig8c.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images 
were produces with 29 kHz bandwidth and 51.2 ms integration. The calibrated image shows 
significant reduction in sidelobe rumble throughout, while retaining the prominent Cyg~A and 
Cas~A sources. The galactic plane is also substantially more evident after calibration. For 
reference, the GSM is shown in the same coordinates and weighted by two factors of the LWA 
primary beam in the right panel. While much of the GSM is visible in the calibrated image, we 
note that the model used to calibrate was only the two bright sources, Cyg~A and Cas~A.
}
\label{fig:data_images}
\end{center}
\end{figure*}

To compare the image qualities quantitatively we compute the dynamic range, defined as the 
peak of the image over the noise level. We estimate the noise level as the median of the 
absolute deviation of the image, a robust statistic. With this metric we find the calibrated image 
to have a 55\% dynamic range improvement over the uncalibrated image.

\section{Noise trends and error analysis}\label{sec:noise}
Now we return to simulations to study the effect of noise in our system, and the consequences 
of an incomplete sky model. We run a suite of simulations varying the receiver noise and 
integration times, while forming antenna-pixel correlations, $\Cna$, for EPICal gain solutions, 
and simultaneously forming visibilities from the same E-field streams to find visibility-based 
gain solutions for comparison.

>>>>

Images created at the beginning of calibration and at the end are shown in 
figure~\ref{fig:sim_images}. Each image is obtained over 10~ms integration, corresponding to 
all snapshot images created with a given set of gain estimates. The top panel shows the image 
produced with our initialized unity gains. Because the phases are completely random, the 
image is essentially noise with the primary beam evident. After 20 iterations, the image is far 
more clear, shown in the bottom panel. Each of the ten simulated sources is clearly visible, 
indicated with red circles.

@@

\section{Application to LWA data}\label{sec:data}
We next demonstrate our calibration algorithm using an observation from the LWA station in 
New Mexico \citep{ell13}. The data is from the LWA narrow-band transient buffer (TBN), with time ordered 
voltage data from the 255 core antennas within a 110~m x 100~m ellipse. 
The central frequency is 74.03 
MHz, with a bandwidth of 100 kHz and sampling period of 5.12 ms (frequency channel 
resolution of 195.3125 Hz). For this demonstration we limit ourselves to a single polarization.

@@

We made several choices while studying the behavior of the LWA data to improve our 
calibration. Through our previous imaging work, we noted that the flux scale of uncalibrated 
images was consistent with average gain amplitudes of 0.25. To allow the calibration to 
converge quickly we initialized our gain estimates at this level. 
We also found a boost in signal to noise is achieved easily by assuming the gains are
constant across a range of frequencies, and averaging solutions across channels.
Here we average solutions across 150 channels, or about 29 kHz. 
With a fractional bandwidth $B/f_0 = 3.9 \times 10^{-4} \ll 1$, we assume a smooth bandpass 
across the band. 
A damping factor of $\gamma = 0.7$ was adopted.
The large damping factor, relative to what was used in simulations, was necessary due to
the larger noise levels in the LWA data compared to the ideal simulations.

Finally, we found that seven antennas\footnote{LWA antenna IDs 48, 85, 124, 148, 203, 217, 
and 244 were omitted.} produced unstable gain solutions, and in fact corrupted the entire array. 
We therefore omitted these antennas from our analysis, resulting in a total of 248 antennas to 
calibrate. In each calibration loop, we form $C_{i,\hat{\boldsymbol{s}}_0}, i=1,2,\ldots N_a$ and 
update our gain estimates over 10 samples (51.2 ms). We iterate the loop 30 times for a 
total of 1.536 seconds. The results of this calibration experiment are shown in figures~
\ref{fig:data_phase}~--~\ref{fig:data_images}.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.3\linewidth]{fig8a.pdf}
\includegraphics[width=0.3\linewidth]{fig8b.pdf}
\includegraphics[width=0.3\linewidth]{fig8c.pdf}
\caption{Images produced before (left) and after (middle) calibrating LWA data. These images 
were produced with 29 kHz bandwidth and 51.2 ms integration. The calibrated image shows 
significant reduction in sidelobe rumble throughout, while retaining the prominent Cyg~A and 
Cas~A sources. The galactic plane is also substantially more evident after calibration. For 
reference, the GSM is shown in the same coordinates and weighted by two factors of the LWA 
primary beam in the right panel. While much of the GSM is visible in the calibrated image, we 
note that the model used to calibrate was only the two bright sources, Cyg~A and Cas~A.
}
\label{fig:data_images}
\end{center}
\end{figure*}

To compare the image qualities quantitatively we compute the dynamic range, defined as the 
peak of the image over the noise level. We estimate the noise level as the median of the 
absolute deviation of the image, a robust statistic. With this metric we find the uncalibrated 
and calibrated images to have dynamic ranges of 88.5 and 137, respectively, a 55\% 
improvement.

\section{Noise trends}\label{sec:noise}
Now we return to simulations to study the effect of noise in our system, and the consequences 
of an incomplete sky model. We run a suite of simulations varying the receiver noise and 
integration times, while forming antenna-pixel correlations, $\Cna$, for EPICal gain solutions, 
and simultaneously forming visibilities from the same E-field streams to find visibility-based 
gain solutions for comparison.
